How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-03-13 docs/manuscript.Rmd
poisson models with overdispersion
updating sections on travel times & bite incidence
- also cleaning up bezier & plotting scripts
- simplified shapefiles for easier plotting and lower file sizes (yay!)

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)

@@

## data
ctar_metadata <- fread("../data/raw/ctar_metadata.csv")
district_bites <- fread("../output/bites/district_bites.csv")
mora_bites <- fread("../output/bites/mora_bites.csv")
mada_communes <- readOGR("../data/processed/shapefiles/mada_communes.shp")
mada_districts <- readOGR("../data/processed/shapefiles/mada_districts.shp")

@@

@@

## Estimating travel time to closest ARMC {#methods-spatial}
We used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale. We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see supplementary section S1](#S1)). To asses the accuracy of this method, we compared this estimate to driving time data collected by IPM during field missions and self-reported travel times from a subset of patients reporting to the Moramanga ARMC. We also compared unweighted average travel times and distance  Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, [cite older WP paper here!]) and aggregated to the commune and district levels. We compared these population estimates to recently released estimates from the 2018 census of Madagascar, and a recent census of three communes in the Moramanga District (see SZ).

## Estimating average annual bite incidence {#methods-bitedata}
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually. These data include details of the administrative district of the bite patient’s address and the date of reporting.

For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted. We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days ([Figure S2.1A and B](#figS2.1)). To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

The 31 months of data available from the Moramanga district resolved to the commune level (the administrative level below the district, Figure 1B) provide a useful comparison to the national scale data. These data were completely reported with details on the type of exposure documented. We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

## Modeling reported bite incidence as a function of travel time to closest ARMC {#methods-bitemods}
We modeled incidence of bites as a function of travel time ($T$) as follows: 
$$ \mu = exp(\beta_{t}T \ + \ \beta_0)P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC, and the human population size, $P$ (an offset which scales the predicted number of bites), for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j\ + \ \beta_0)P_j$$

As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel time. For the models with population size, we used a model framework with either population size alone ($\mu = exp(\beta_pP + \beta_0)$) or population size and travel times ($\mu = exp(\beta_{t}T + \beta_pP + \beta_0)$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

@@

Finally, to check how correcting for underreporting affected our modeling results, we fitted these same models to data uncorrected for reporting (see [supplementary section S4](#S4)).

@@

```{r fig 2, fig.fullwidth=TRUE}
include_graphics("../figs/main/M2.jpeg")
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

@@

```{r ncsbs}
## candidate points
csbs <- read.csv("../data/raw/csbs.csv", stringsAsFactors = FALSE)
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

@@

## Estimating travel times to and catchment populations of 
Travel times to the closest ARMC varies greatly across the country. We compared these estimates based
on the MAP friction surface to self-reported estimates of hours traveled to the Moramanga clinic for a subset of patients and driving times across the country from IPM. We found that the travel times estimated from the friction surface underestimate true driving times and patient reported travel times to the Moramanga ARMC. Importantly, the friction surface does not account for road condition, traffic, or seasonality in travel times. It also assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence. From the self reported travel patterns, we know that people use mutliple modes of transport and walking is common adding to the variation (SX). Despite these limitations, the estimated travel times correlate strongly with the reported and groundtruthed driving times. Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (SY). We use the estimated weighted mean of travel times by population as a proxy for access at the commune/district level in subsequent analyses.

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1.2). Therefore, we defined the catchment for each clinic as all districts for which the clinic was the closest ARMC for the majority of the population in that district. We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 1A).

```{r fig 1, fig.fullwidth=TRUE}
include_graphics("../figs/main/M1.jpeg")
```
*__Figure 1.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data. Color corresponds to the clinic catchment. Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients. Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. __B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district. Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic. __C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis). Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e. the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*

## Estimating average annual bite incidence at the district and commune level
We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics ([Fig S2.1C](#figS2.1)). See [supplementary section S2](#S2) for more details.

Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C). Scale discussion. Outlier discussion.

## Modeling reported bite incidence as a function of travel times {#results-bitemods}
All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2). Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).  Sensitivity to underreproting. For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2). In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3). These predictions were robust to how we corrected for underreporting ([supplementary section S4](#S4)). As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence.

```{r fig 3, fig.fullwidth=TRUE}
include_graphics("../figs/main/M3.jpeg")
```
_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis). The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept. The envelopes shows the 95% CI given the random intercept for the commune and district models. The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept. The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model)._

@@

```{r burden results}
natl_burden <- read.csv("../output/preds/natl_burden.csv")
```

@@

```{r fig 4, fig.fullwidth=TRUE}
include_graphics("../figs/main/M4.jpeg")
```
*__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons. __A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district. Points are colored by and districts are ordered by travel times. The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models. Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current CTAR.*

## The impact of expanding PEP access to additional clinics {#results-incremental}
 We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 100 {inline} clinics reducing burden by almost 50% {inline}. However, benefits accrue less rapidly as ARMC are added, saturating after approximately 200 {inline} additional clinics. As clinics are added, reported bite incidence saturates, and instead there are shifts in which clinics bites are reported to ([see supplementary section S5](#S5)). Even at this maximal PEP access, our model still predicts > 100 {inline} rabies deaths per annum, as average reporting of rabid exposures still remains at approximately 85 % ([Fig S5.4](#figS5.4)). This is not due to saturation of reporting (see distribution of reporting fig in S5), but instead to to broad patterns of remoteness across Mada. Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded. Our model predictes an increase from 40,000 {inline} vials under current clinic provisioning, to ~50,000 {inline} vials with 200 {inline} clinics, ~60,000 {inline} and approximately 90,000 {inline} vials if all clinics were to provision PEP.

@@

```{r fig 5, fig.fullwidth=TRUE}
include_graphics("../figs/main/M5.jpeg")
```

@@

## Limitations
Our modeling framework makes strong assumptions about baseline rabies exposure incidence, assuming a range of 11 - 55 exposures per 100,000 persons in Madagascar. We base these estimates on a reasonable range of human:dog ratios [@knobelReevaluatingBurdenRabies2005; @ratsitorahinaDogEcologyDemography2009; @sambo2018estimating] and dog rabies incidence level [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. A previous study in the Moramanga District in Madagascar estimated a range of 42 – 110 exposures per 100,000 persons, however this was over a 16 month period with frequent rabies cases and multiple human deaths [@rajeevHealthcareUtilizationProvisioning2018]. As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we sample uniformly over this range and over the proportion of reported bites estimated to be rabid (0.2 - 0.6 from the Moramanga data) to generate a mean estimate of burden. Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same. While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies incidence. Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.

The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported). Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems. 
<!-- TO DO: SOMETHING ABT SHORTAGES HERE AND BITS FROM CLAIRE'S MSC -->

We make several simplifying assumptions regarding PEP provisioning and efficacy. First, we assume perfect compliance of patients and efficacy of PEP. Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018]. We also assume that patients report to their closest clinic, but in our data approximately 10% of patients reported to other clinics. As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level. Current practices of provisioning PEP reactively to Category I exposures. Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped [@rajeevHealthcareUtilizationProvisioning2018].

We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008]. Importantly, in the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel. In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that interact to determine an individual’s access to PEP. While we did not explicitly include variation in clinic functioning, we did include a catchment level random intercept, to account for spatial differences in reported bite incidence not due to travel times. This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e. differences in clinic functioning such as vaccine availability) and at the catchment level (i.e. differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.

## Broader context
```{r}
burden_preds <- fread("../output/preds/complete/burden_filled.csv")
## Grouped to district
burden_preds$distcode <- mada_communes$distcode[burden_preds$names]
burden_preds$commcode <- mada_communes$commcode[burden_preds$names]

>>>>

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)

@@

## data
ctar_metadata <- fread(here("data/raw/ctar_metadata.csv"))
district_bites <- fread(here("output/bites/district_bites.csv"))
mora_bites <- fread(here("output/bites/mora_bites.csv"))
mada_communes <- readOGR(here("data/processed/shapefiles/mada_communes.shp"))
mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))

@@

# AUTHOR SUMMARY

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale. We then compared three different metrics of access:

1) Travel time
2) Weighted travel time
3) Distance

extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

We compared these travel time and distance estimates to driving time data collected by IPM during field missions and patient reported travel times from a subset of patients reporting to the Moramanga ARMC (see Fig S1.2 for raw data).

## Estimating average annual bite incidence {#methods-bitedata}
We used two datasets on bite patients reporting to ARMC:
1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually that included data on the date of reporting and were resolved to the district level (i.e. the district where the patient resided)
2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level (the administrative level below the district). We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for this, we estimated clinic level undersubmission as the proportion of days on which forms were submitted in a given year, excluding any periods for which there were no forms submitted for 15 consecutive days. We then divided the total number of bites reported to each district at each clinic in a given year by the undersubmission rate for that clinic in that year. Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates. The data from the Moramanga district were completely reported across the 31 month period.

At both the commune and district scale, we assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit. For the national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) as follows: 
$$ \mu = e^{(\beta_{t}T \ + \ \beta_0 \ )}P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T$) and the human population size ($P$) (an offset which scales the predicted number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu = \sum \limits_{j=1}^je^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$

As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel times. For the models with population size, we used a model framework with either population size alone ($\mu = e^{(\beta_pP + \beta_0 )}$) or population size and travel times ($\mu = e^{(\beta_{t}T + \beta_pP + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

@@

Since we're assuming for poisson errors: we tested for overdispersion.

Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to uncorrected data.

@@

```{r fig 2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M2.jpeg"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/raw/csbs.csv", stringsAsFactors = FALSE))
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

@@

## Estimating travel times to the closest ARMC
```{r ttime stats}
ttime_stats <- read.csv(here("output/stats/prop_pop_ttimes.csv"))
# get prop of pop > 3 hrs away at baseline
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
  sum(prop_pop) -> prop_over3_base
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
  sum(prop_pop) -> prop_over6_base
```

Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population living greater than 3 hrs away from a clinic (Fig 3B). We found that travel times estimated from the friction surface generally underestimate true driving times and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2). The friction surface assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence (i.e. if there is a road, then people will travel at vehicle rather than walking speeds). However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3). Despite these discrepancies, travel times correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4). Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1). Therefore we use this metric as a proxy for access at the commune/district level in subsequent analyses.

```{r fig M2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M3_ttimes.jpeg"))
```
*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar. __B)__ Distribution of the population across travel times. __C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level. For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune. The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown. The dashed black line is the 1:1 line.*

## Estimating average annual bite incidence at the district and commune level
We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population. While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B). In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A) and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).

```{r fig M3 catchmaps}
include_graphics(here("figs/main/M3_bitemaps.jpeg"))
```
*__Figure 3.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each ARMC for which we have data. Color corresponds to the clinic catchment. Circles with white outline are the total number of bites reported for that district. Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded. Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. Out-of-catchment reporting is indicated where points and line colors are mismatched. __B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the area plotted. The black outline shows the district of Moramanga, and the red polygons shows all the communes included in the Moramanga ARMC catchment. Circle with black outline is the location of the Moramanga ARMC with the size of points proportional to the total number of bites reported to the clinic, as per A.*

Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1). However, we found that corrected estimates of reproted bites generated predictions of vial demand close to what was provisioned over the same time period ([Figure S2.2A and B](#figS2.2)). We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 3A). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).

```{r fig M4 biteinc}
include_graphics(here("figs/main/M4_biteinc.jpeg"))
```

*__Figure 4.__ The estimated average annual bite incidence from the national and Moramanga data plotted at the __A)__ district scale (both datasets) and at the __B)__ commune scale (Moramanga dataset). Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e. the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used). The point lines indicate the range of estimated bite incidence for each district.*

```{r}
outliers <- read.csv(here("output/stats/outliers.csv"))
head(outliers)
```

Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship. There were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher incidence at low travel times at the commune level compared to the district.

## Modeling reported bite incidence as a function of travel times {#results-bitemods}

All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2). Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3). Sensitivity to underreporting. For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2). In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3). These predictions were robust to how we corrected for incomplete submission of data by clinics ([supplementary section S4](#S4)). As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence. (OD COMP--use the Moramanga model with OD & the District model with OD)

```{r fig 3, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M3.jpeg"))
```
_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis). The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept. The envelopes shows the 95% CI given the random intercept for the commune and district models. The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept. The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model)._

@@

```{r burden results}
natl_burden <- read.csv(here("output/preds/natl_burden.csv"))
```

@@

```{r fig 4, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M4.jpeg"))
```
*__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons. __A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district. Points are colored by and districts are ordered by travel times. The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models. Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current CTAR.*

## The impact of expanding PEP access to additional clinics {#results-incremental}

 We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 100 {inline} clinics reducing burden by almost 50% {inline}. However, benefits accrue less rapidly as ARMC are added, saturating after approximately 200 {inline} additional clinics. As clinics are added, reported bite incidence saturates, and instead there are shifts in which clinics bites are reported to ([see supplementary section S5](#S5)). Even at this maximal PEP access, our model still predicts > 100 {inline} rabies deaths per annum, as average reporting of rabid exposures still remains at approximately 85 % ([Fig S5.4](#figS5.4)). This is not due to saturation of reporting (see distribution of reporting fig in S5), but instead to to broad patterns of remoteness across Mada. Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded. Our model predictes an increase from 40,000 {inline} vials under current clinic provisioning, to ~50,000 {inline} vials with 200 {inline} clinics, ~60,000 {inline} and approximately 90,000 {inline} vials if all clinics were to provision PEP.

@@

```{r fig 5, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M5.jpeg"))
```

@@

## Limitations
We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008]. In the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel. In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.

Importantly,the travel time estimates from the MAP friction surface underestimate true travel times reported by patients. This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018]. In Madagascar, the presence of paved roads does not necessarily reflect their quality and the modes of transport used to travel these roads. Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune. Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access.

<!-- TO DO: UPDATE TO REFLECT CHANGES IN OD MODS ETC. -->

We did include a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times. This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e. differences in clinic functioning such as vaccine availability) and at the catchment level (i.e. differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.

Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior. We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning.

<!-- TO DO: CLAIRE'S MSC ON OTHER REASONS + CLINIC FUNCTIONING -->
<!-- TO DO: OUTLIERS...SCALE? -->

<!-- TO DO: UPDATE BASED ON CHANGES TO BURDEN MODEL ETC. -->
Our modeling framework makes strong assumptions about baseline rabies exposure incidence, assuming a range of 11 - 55 exposures per 100,000 persons in Madagascar. We base these estimates on a reasonable range of human:dog ratios [@knobelReevaluatingBurdenRabies2005; @ratsitorahinaDogEcologyDemography2009; @sambo2018estimating] and dog rabies incidence level [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. A previous study in the Moramanga District in Madagascar estimated a range of 42 – 110 exposures per 100,000 persons, however this was over a 16 month period with frequent rabies cases and multiple human deaths [@rajeevHealthcareUtilizationProvisioning2018]. As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we sample uniformly over this range and over the proportion of reported bites estimated to be rabid (0.2 - 0.6 from the Moramanga data) to generate a mean estimate of burden. Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same. While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies incidence. Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.

The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported). Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.

We make several simplifying assumptions regarding PEP provisioning and efficacy. First, we assume perfect compliance of patients and efficacy of PEP. Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018]. We also assume that patients report to their closest clinic, but in our data approximately 10% of patients reported to other clinics. As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level.

<!-- TO DO: CAT 1 -->

Current practices of provisioning PEP reactively to Category I exposures. Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped [@rajeevHealthcareUtilizationProvisioning2018].
